%==============================================================================
% Paper A -- skeleton (bullets, to be polished into prose)
% Target venue: systems / measurement (e.g., a virtualization or measurement
%   workshop). Reviewers: the two IEEE professors flagged this as the strongest
%   standalone systems contribution.
%
% Build: pdflatex skeleton.tex   (needs IEEEtran; if unavailable, change the
%   \documentclass line below to \documentclass[11pt]{article} and delete the
%   \IEEEauthorblock* macros.)
%==============================================================================
\documentclass[conference]{IEEEtran}
\usepackage{amsmath,booktabs,siunitx,url}

\title{Active Page Fraction: A Single-Scalar Liveness Signal for\\
Whole-VM Memory Introspection at Tractable Cost}

\author{\IEEEauthorblockN{TODO Author list}
\IEEEauthorblockA{TODO Affiliation\\
\texttt{TODO@email}}}

\begin{document}
\maketitle

%------------------------------------------------------------------
\begin{abstract}
% TODO: polish into 150-200 words of prose. Bullet stub:
\begin{itemize}
  \item Problem: per-page, two-dimensional memory-change signals (cosine and
        Hamming over every 4\,KiB page) are information-rich but storage- and
        compute-prohibitive for repeated whole-VM capture.
  \item Idea: collapse the per-snapshot signal to a single scalar, the
        \emph{Active Page Fraction} (APF) -- the fraction of guest pages whose
        bytes change between consecutive full-memory snapshots.
  \item Mechanism: a streaming capture architecture (``B+3.1'') in which an
        asynchronous helper computes the diff and, in its delete-as-you-go
        mode, removes the prior dump to bound transient disk, while persisting
        only a 1-D trajectory. (The v3 run retained dumps; see
        Sec.~\ref{sec:results}.)
  \item Result: an estimated reduction of order $10^{6}$ in persistent
        storage for a fixed capture campaign (a projection; see
        Sec.~\ref{sec:results}), with the dominant capture cost set by the
        hypervisor snapshot (\texttt{pmemsave}) rather than by the chosen
        sampling interval.
\end{itemize}
\end{abstract}

%------------------------------------------------------------------
\section{Introduction}
% Proposed introduction (bullets):
\begin{itemize}
  \item Hypervisor-external memory introspection observes a guest without an
        in-guest agent; it is attractive for security monitoring and workload
        characterization because it is hard for the guest to evade.
  \item A natural raw signal is the per-page change between successive
        full-memory snapshots. Prior in-project work (``Pipeline A'') computed,
        for each 4\,KiB page, both a cosine and a Hamming similarity, yielding a
        two-dimensional matrix of shape $[T \text{ snapshots} \times N \text{ pages}]$
        with $N \approx 262{,}144$ pages for a 1\,GiB guest.
  \item This 2-D signal is faithful but heavy: retaining the dumps needed to
        recompute it costs on the order of \SI{60}{\gibi\byte} per long cell and
        a projected $\approx$\SI{5}{\tebi\byte} for a 90-cell campaign (never
        fully retained) -- infeasible on the available host.
  \item \textbf{Gap}: a memory-liveness signal that is cheap enough to capture
        repeatedly across many workloads and durations, while preserving enough
        structure to (a) distinguish active from idle phases and (b) feed
        downstream tuning, classification, and segmentation.
  \item \textbf{This paper} proposes the Active Page Fraction (APF): one scalar
        per snapshot-pair, defined as the fraction of guest pages that changed
        at all. We present a streaming capture architecture that makes APF
        capture disk-bounded, and we characterize the true cost floor of the
        method.
  \item \textbf{Contributions} (expanded in Sec.~\ref{sec:contrib}):
    \begin{itemize}
      \item C1: the APF signal and its streaming, delete-as-you-go capture
            pipeline;
      \item C2: a measured cost model showing the dominant cost is the
            hypervisor snapshot (suspend/\texttt{pmemsave}/resume), not the
            requested sampling interval;
      \item C3: an open, reproducible 132-cell dataset and the analysis tooling
            built on it.
    \end{itemize}
\end{itemize}

%------------------------------------------------------------------
\section{Background and Related Work}
\begin{itemize}
  \item VM introspection and memory forensics \cite{garfinkel2003vmi}: external
        observation of guest memory. TODO: position against agentless
        monitoring and live-migration dirty-page tracking.
  \item Working-set estimation / dirty-bit sampling \cite{TODO_working_set_estimation}:
        hypervisors already track dirtied pages for migration; APF differs by
        being content-derived (byte difference) and capture-external rather than
        relying on hardware dirty bits.
  \item Memory deduplication / same-page merging \cite{TODO_ksm_dedup}:
        content-similarity at page granularity; contrast with APF's coarser
        ``changed-or-not'' bit.
  \item TODO: cite snapshotting/checkpointing cost studies to frame the
        \texttt{pmemsave} overhead.
\end{itemize}

%------------------------------------------------------------------
\section{The Active Page Fraction Signal}
\begin{itemize}
  \item \textbf{Capture primitive.} Per snapshot cycle the producer issues
        \texttt{virsh suspend}, waits for \texttt{domstate==paused}, invokes the
        QEMU monitor \texttt{pmemsave(val:0, size:ramBytes, file)} to write a
        flat RAW image of all guest physical memory, then \texttt{virsh resume}.
        Guest RAM is \SI{1024}{\mebi\byte}; each dump is exactly \SI{1}{\gibi\byte};
        page size is \SI{4096}{\byte} ($N = 262{,}144$ pages).
  \item \textbf{APF definition.} For consecutive dumps $a$, $b$ mapped as
        \texttt{uint8} via \texttt{numpy.memmap} and reshaped to
        $[N, 4096]$:
        \[
          \mathrm{APF} \;=\; \frac{1}{N}\sum_{p=1}^{N}
          \mathbf{1}\!\left[\exists j: a_{p,j} \neq b_{p,j}\right].
        \]
        That is, a page contributes 1 if any byte differs. A single-bit flip and
        a full-page rewrite are weighted equally (a deliberate coarsening).
  \item \textbf{Streaming architecture (B+3.1).} An asynchronous helper
        (\texttt{ionice -c3 nice -n19}) computes each pair's APF, appends one
        JSONL record via an atomic \texttt{O\_APPEND} write, emits an ack, and
        -- in its delete-as-you-go mode -- \emph{deletes the older dump}, so at
        most a small number of dumps coexist and transient disk is capped at
        roughly \SI{2}{}--\SI{3}{\gibi\byte}. \emph{Honesty note:} the v3
        capture ran with \texttt{keep\_dumps} enabled, which \emph{disables} the
        delete step. The helper still computed APF in stream, but every dump was
        retained until a single end-of-cell sweep (cell notes record, e.g.,
        ``removed 29 dumps''), so the \SI{2}{}--\SI{3}{\gibi\byte} cap is a
        design target \emph{not} exercised in v3; measured peak transient disk
        scaled with trajectory length (up to one \SI{1}{\gibi\byte} dump per
        snapshot, i.e.\ tens of \si{\gibi\byte} for a 600\,s cell).
  \item \textbf{Barrier.} A cell-end barrier waits for outstanding helper acks
        (hard cap \SI{180}{\second}, \SI{15}{\second} idle quit) before metrics
        are computed, guaranteeing the trajectory is complete.
  \item TODO: figure -- timeline of suspend/pmemsave/resume overlapped with the
        async diff+delete helper.
\end{itemize}

%------------------------------------------------------------------
\section{Contributions in Detail}
\label{sec:contrib}
\begin{itemize}
  \item \textbf{C1 -- signal + pipeline}: APF reduces the per-snapshot signal
        from a length-$N$ vector (2-D over time) to a single scalar (1-D over
        time), and the streaming helper makes its capture disk-bounded.
  \item \textbf{C2 -- cost model}: we show the effective cadence is set by the
        snapshot primitive, not the configured interval (Sec.~\ref{sec:results}).
  \item \textbf{C3 -- dataset}: 132 cells spanning 11 workloads $\times$ 3
        durations $\times$ 4 replicates, captured at a single requested interval,
        with the full tooling released.
\end{itemize}

%------------------------------------------------------------------
\section{Evaluation}
\label{sec:results}
% Proposed results (bullets with REAL numbers from the v3 run):
\begin{itemize}
  \item \textbf{Storage reduction.} For a 90-cell campaign the APF trajectory
        persists $\approx$\SI{4.7}{\mebi\byte}; retaining the dumps needed for
        the 2-D per-page pipeline would project to $\approx$\SI{5}{\tebi\byte}
        -- an estimated ratio of order $10^{6}$ (projection, not a measured
        numerator). In v3 the delete step was disabled
        (\texttt{keep\_dumps}=true), so transient disk instead scaled with
        trajectory length rather than holding at the
        \SI{2}{}--\SI{3}{\gibi\byte} design target; validating that cap requires
        a delete-as-you-go re-run.
        % HONESTY (reviewer A): the 5 TiB numerator is a PROJECTION for a
        % campaign never fully retained, and 4.7 MiB is the 90-cell figure
        % while the dataset is 132 cells. Before publication: reconcile to one
        % campaign size, report a MEASURED numerator (or label it a projection),
        % and present the ratio as an estimate -- do not lead with a bare
        % "10^6x" superlative.
  \item \textbf{Compute reduction.} Per-pair APF via \texttt{memmap} byte-diff
        takes \SI{300}{}--\SI{500}{\milli\second}, versus \SI{1.5}{}--\SI{3}{\second}
        for the Rust per-page cosine+Hamming kernel -- lower host contention and
        more snapshots per cell. (This pair is not separately logged in v3;
        before publication, measure it under controlled cache state.)
  \item \textbf{Cost is the snapshot, and it is variable.} Although the
        requested guest interval was \SI{500}{\milli\second} (and the guest
        indeed ran in $\approx$\SI{0.52}{\second} bursts between snapshots), the
        host snapshot cycle is far larger and \emph{not} a fixed floor: median
        \SI{9.75}{\second} per snapshot, range \SI{2.0}{}--\SI{23.6}{\second}
        over the 128 \texttt{ok} cells. Snapshot counts per cell are
        correspondingly spread -- 120\,s: median 14 (range 5--43); 300\,s:
        median 26 (12--148); 600\,s: median 53 (28--100) -- so the often-quoted
        ``7/20/28'' are near the per-duration \emph{minima}, not typical values.
        The cost is dominated by \texttt{pmemsave}: median \SI{7.06}{\second}, a
        median 81\% of the cycle, with suspend ($\approx$\SI{0.35}{\second}) and
        resume ($\approx$\SI{0.05}{\second}) negligible and the asynchronous
        helper read-back off the critical path. The estimated VM-pause fraction
        is high but not uniform: median 0.946 (range 0.74--0.98).
  \item \textbf{Implication.} Because the host snapshot cycle dwarfs the
        \SI{500}{\milli\second} interval -- the guest already idles
        $\approx$\SI{0.52}{\second} between snapshots -- tuning the interval
        below the cycle is essentially a no-op. This motivates Paper B's
        single-interval design and reframes ``how often to sample'' as ``how
        cheaply can we snapshot,'' i.e.\ how to shrink \texttt{pmemsave}.
  \item \textbf{Signal fidelity (forward reference).} Despite the coarsening,
        APF retains a weak but measurable family signal (Paper C, after
        controlling for feature-construction leakage) and steady workloads are
        stationary under windowed analysis (Paper D). We claim no instance-level
        fidelity.
  \item TODO: table -- per-duration (snapshots, mean APF, pause fraction);
        figure -- example APF trajectories for one phasic and one steady cell.
\end{itemize}

%------------------------------------------------------------------
\section{Limitations}
\begin{itemize}
  \item APF is intentionally coarse: a 1-bit change and a full rewrite are
        indistinguishable. Per-page sub-signatures require re-introducing the
        2-D pipeline (deferred future work).
  \item The cadence (median $\approx$\SI{10}{\second}/snap, up to
        $\approx$\SI{24}{\second}) limits temporal resolution; trajectories are
        short (median 30 points, range 4--99), which constrains spectral
        analysis (see Paper B's coverage limitation).
  \item \texttt{pmemsave} suspends the guest (median $\approx$95\% pause); the
        method as implemented is an offline/research capture, not a
        production-latency introspector. Live dirty-bit tracking is the natural
        optimization.
\end{itemize}

%------------------------------------------------------------------
\section{Conclusion}
\begin{itemize}
  \item APF turns whole-VM memory-change observation from a multi-terabyte,
        per-page exercise into a megabyte-scale single-scalar trajectory.
  \item The binding cost is the hypervisor snapshot (\texttt{pmemsave}), which
        both bounds temporal resolution and frees the sampling-interval knob --
        a finding that shapes the rest of the portfolio.
\end{itemize}

\bibliographystyle{IEEEtran}
\bibliography{../refs}
\end{document}
